Modelling BCG vaccination in the UK

### 2019-08-06 rmds/01-introduction.Rmd
update word count
intro reviewed

@@

The availability of data is revolutionising the way we view the world; in few other areas has this revolution been felt more than in public health. In 2000, Public Health England (PHE) launched a routine surveillance system for TB, which records demographic, clinical, and microbiological information on all notified cases.[@PHE2017] This data-set allows us to study the details of TB epidemiology in England more easily than ever before. Whilst this information would present much of interest by itself, by combining it with other data-sets we can adjust for the changing demographics of the English population to study the trends in TB over time.

This thesis explores the impact of changing BCG policy in England, with the aim of informing global efforts to control TB. As a first step I outline some of the key background information and discuss the tooling that I developed in order to explore this information more fully. I then make use of the detailed PHE routine TB surveillance data to explore the current epidemiology of TB in England. Next, I use statistical models that make use of this data to explore the impact from the 2005 change in BCG policy. Finally I develop, and fit, a detailed, semi-sthochastic, mechanistic model of TB and BCG vaccination in England in order to forecast the ongoing impact of the change in policy versus multiple alternative scenarios. This thesis is also available as a website^[Website: https://www.samabbott.co.uk/thesis/], a pdf^[PDF: https://www.samabbott.co.uk/thesis/thesis.pdf], and a reproducible Rmarkdown^[Rmarkdown: https://github.com/seabbs/thesis]

@@

In this thesis, a partially observed semi-stochastic compartmental model of TB is developed that models demographic processes such as ageing, births and deaths, as well as vaccination and TB treatment (Chapter \@ref(model-development)). Compartmental infectious disease models generally operate by separating a given population into a series of groups, most commonly susceptible, infectious and recovered populations.[@Anderson1991; @Keeling2007] Movements between these groups are then modeled using a series of differential equations. Transmission is modeled using mass action,[@Anderson1991; @Keeling2007] where infected cases are assumed to randomly interact with susceptible individuals. Additional detail can be added to this model by stratifying the population further and adding additional parameters to modify the degree of mixing between populations. Transition between compartments is assumed to be exponential. See [@Anderson1991; @Keeling2007] for a theoretical introduction to infectious disease models and [@King] for implementation details using R.

@@

* **Chapter \@ref(data):** This chapter describes the epidemiology of TB in England, using routine national data-sets. Focusing on: the impact of missing data; the mechanisms underlying that missing date; seasonal trends; the role of age; UK birth status; BCG status; trends in TB incidence over time; and TB outcomes in England using case rates. These data are used in all subsequent chapters in this thesis.

@@

* **Chapter \@ref(direct-eff):** This chapter uses multilevel statistical models to assess the effects of the 2005 change in BCG vaccination policy on the populations targeted by each vaccination scheme.

@@

* **Chapter \@ref(model-results):** In this chapter the model, developed and fitted in the previous chapters, is used to forecast the impact of universal BCG vaccination at school-age vs. universal vaccination of neonates vs. no vaccination from 2005 on-wards. The results from this forecasting are compared to the actual notifications recorded from 2005 to 2015, with the divergence from the observed data discussed. The ongoing impact of each policy is then discussed through to 2030.

>>>>

```{r, include = FALSE}
## Wordcount: 2934
```

@@

The availability of data is revolutionising the way we view the world; in few other areas has this revolution been felt more than in public health. In 2000, Public Health England (PHE) launched a routine surveillance system for TB, which records demographic, clinical, and microbiological information on all notified cases.[@PHE2017] This dataset allows us to study the details of TB epidemiology in England more easily than ever before. Whilst this information would present much of interest by itself, by combining it with other datasets we can adjust for the changing demographics of the English population to study the trends in TB over time.

This thesis explores the impact of changing BCG policy in England, with the aim of informing global efforts to control TB. As a first step I outline some of the key background information and discuss the tooling that I developed in order to explore this information more fully. I then make use of the detailed PHE routine TB surveillance data to explore the current epidemiology of TB in England. Next, I use statistical models that make use of this data to explore the impact from the 2005 change in BCG policy. Finally I develop, and fit, a detailed, semi-sthochastic, mechanistic model of TB and BCG vaccination in England in order to forecast the ongoing impact of the change in policy versus multiple alternative scenarios. This thesis is also available as a website^[Website: https://www.samabbott.co.uk/thesis/], a pdf^[PDF: https://www.samabbott.co.uk/thesis/thesis.pdf], and a reproducible Rmarkdown document^[Rmarkdown: https://github.com/seabbs/thesis]

@@

In this thesis, a partially observed semi-stochastic compartmental model of TB is developed that models demographic processes such as ageing, births and deaths, as well as vaccination and TB treatment (Chapter \@ref(model-development)). Compartmental infectious disease models generally operate by separating a given population into a series of groups, most commonly susceptible, infectious and recovered populations.[@Anderson1991; @Keeling2007] Movements between these groups are then modeled using a series of differential equations. Transmission is modeled using mass action,[@Anderson1991; @Keeling2007] where infected cases are assumed to randomly interact with susceptible individuals at a rate dictated by the concentration of susceptibles in the population. Additional detail can be added to this model by stratifying the population further and adding additional parameters to modify the degree of mixing between populations. Transition between compartments is assumed to be exponential. See [@Anderson1991; @Keeling2007] for a theoretical introduction to infectious disease models and [@King] for implementation details using R.

@@

* **Chapter \@ref(data):** This chapter describes the epidemiology of TB in England, using routine surveillance datasets. Focusing on: the impact of missing data; the mechanisms underlying that missing date; seasonal trends; the role of age; UK birth status; BCG status; trends in TB incidence rates over time; and TB outcomes in England using case rates. These data are used in all subsequent chapters in this thesis.

@@

* **Chapter \@ref(direct-eff):** This chapter uses a series of multilevel statistical models to assess the effects of the 2005 change in BCG vaccination policy on the populations targeted by each vaccination scheme.

@@

* **Chapter \@ref(model-results):** In this chapter the model, developed and fitted in the previous chapters, is used to forecast the impact of universal BCG vaccination at school-age vs. universal vaccination of neonates vs. no vaccination from 2005 on-wards. The ongoing impact of each policy is then discussed through to 2040.
